### 2022-08-15 description2.tex
Fix typo
Add lab comments

% Summarize pair-HMMs. Brief, comment typically used but move onto FST right away
Statistical alignment is typically performed using pairwise hidden Markov
models (pair-HMMs).
% This computational machines consist of two output tapes and a set of states that
% emit symbols onto one or both tapes.
Pair-HMMs have the ability to rigorously model molecular sequence evolution and
can calculate the probability that two sequences are related, represented
$P(X, Y)$ \parencite{yoon_2009_hmm}.
However, a limitation of pair-HMMs is the ability to only model evolution of two
related sequences from an unknown ancestor.
Finite-state transducers (FSTs) have the ability to calculate the probability
that a sequence $Y$ evolved from sequence $X$, represented $P(Y | X)$.
FSTs share similar computational benefits as pair-HMMs in addition to well
established algorithms for combining them in different ways
\parencite{bradley2007transducers}.
A powerful and versatile algorithm is composition, which consists of sending the
output one FST as the input of a second FST.
The FST model implemented in COATi is designed by composing smaller FSTs, each
representing a specific process.

@@

\begin{figure}[h!]
\begin{framed}
\centering
    \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
    \caption{The Evolution FST is assembled by composing a substitution FST and
    an indel FST. Each node represents a state in an FST while arcs display
    possible transitions between states (and their weights). Absorption and
    emission of symbols occurs between states. (a) The substitution FST
    encodes a 64x64 codon substitution model with 64 arcs from M to S. (b)
    The indel FST allows for insertions (I to U) and deletions (D to V).
    Contiguous insertions and deletions are always arranged for insertions to
    precede deletions to limit equivalent alignments.}
    \label{fig:evolution-fst}
\end{framed}
\end{figure}

% FST implementation
% A path through the FST that \green{results} from composing both sequences with
% the Evolution FST represent a pairwise alignment.
% To find the most \green{probable/likely} alignment we must find the shortest
% path through.
The alignment FST is the result from composing both sequences with the Evolution
FST.
Any path through the alignment FST represents a pairwise alignment, while the
shortest path corresponds to the best alignment.
All FST operations including creation of models, composition, search for the
shortest path, and other optimization algorithms are performed using the C++
openFST library \parencite{allauzen2007openfst}.

>>>>

% Summarize pair-HMMs. Brief, comment typically used but move onto FST right away
Statistical alignment is typically performed using pairwise hidden Markov
models (pair-HMMs).
% This computational machines consist of two output tapes and a set of states that
% emit symbols onto one or both tapes.
Pair-HMMs have the ability to rigorously model molecular sequence evolution and
can calculate the probability that two sequences are related, represented
$P(X, Y)$ \parencite{yoon_2009_hmm}.
However, a limitation of pair-HMMs is the ability to only model the evolution of
two related sequences from an unknown ancestor.
Finite-state transducers (FSTs) have the ability to calculate the probability
that a sequence $Y$ evolved from sequence $X$, represented $P(Y | X)$.
FSTs have similar benefits to pair-HMMs in addition to
well-established algorithms for combining FSTs in different ways
\parencite{bradley2007transducers}.
A powerful and versatile algorithm is composition, which consists of sending the
output of one FST into the input of a second FST.
The FST model implemented in COATi is designed by composing smaller FSTs, each
representing a specific process.

@@

\begin{figure}[h!]
\begin{framed}
\centering
    \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
    \caption{The Evolution FST is assembled by composing a substitution FST and
    an indel FST. Each node represents a state in an FST while arcs display
    possible transitions between states (and their weights). (a) The
    substitution FST encodes a 64x64 codon substitution model with 64 arcs from
    M to S. (b) The indel FST allows for insertions (I to U) and deletions
    (D to V). Contiguous insertions and deletions are always arranged for
    insertions to precede deletions to limit equivalent alignments.}
    \label{fig:evolution-fst}
\end{framed}
\end{figure}

% FST implementation
% A path through the FST that \green{results} from composing both sequences with
% the Evolution FST represent a pairwise alignment.
% To find the most \green{probable/likely} alignment we must find the shortest
% path through.
The alignment FST is the result from composing both sequences with the Evolution
FST.
Any path through the alignment FST represents a pairwise alignment, while the
shortest path corresponds to the best alignment.
All FST operations including model development, composition, search for the
shortest path, and other optimization algorithms are performed using the C++
openFST library \parencite{allauzen2007openfst}.

### 2022-08-15 introduction.tex
Add lab comments

% sequence alignment is important and heavily used
Sequence alignment is a fundamental task in bioinformatics and a cornerstone
step in comparative and functional genomic studies
\parencite{sequence_alignment_rosenberg_2009}.
% often seen as an ad hoc problem
Modern sequence analysis began with the heuristic homology algorithms of
Needleman and Wunsch in 1970 \parencite{identification_smith_1981} and has
progressed to arrive at current aligners.
However, the alignment of molecular sequences is, in practice, often seen as a
tool and the alignment inference as an ad hoc problem
\parencite{morrison_MSA_2018}.

% despite available aligners, there is room for improvement
% summarize sentence into ~ "alignment is performed in the amino acid space"
A common strategy is to perform alignment inference in the amino acid space
\parencite{bininda2005transalign,abascal2010translatorx}.
While this approach is an improvement over DNA models, it discards information,
underperforms compared to alignment at the codon level, and fails in the
presence of artifacts.
% to correct this common practice is to discard lots of data
Given that errors and artifacts are common in molecular data, this requires
costly curation practices that discard large amounts of information.
% errors in genomic data that lead to erroneous downstream analyses
When uncorrected, errors in genomic datasets can lead to erroneous results in
functional and comparative genomic studies \parencite{estimates_schneider_2009}.

>>>>

% sequence alignment is important and heavily used
Sequence alignment is a fundamental task in bioinformatics and a cornerstone
step in comparative and functional genomic studies
\parencite{sequence_alignment_rosenberg_2009}.
% often seen as an ad hoc problem
Modern sequence analysis began with the heuristic homology algorithms of
Needleman and Wunsch in 1970 \parencite{identification_smith_1981} and has
progressed to arrive at current aligners such as Clustal$\Omega$
\parencite{clustal_omega_sievers_2011}, MAFFT \parencite{mafft_katoh_2002}, or
PRANK \parencite{prank_loytynoja_2014}.
However, the alignment of molecular sequences is, in practice, often seen as a
tool and the alignment inference as an ad hoc problem
\parencite{morrison_MSA_2018}.

% despite available aligners, there is room for improvement
% summarize sentence into ~ "alignment is performed in the amino acid space"
A common strategy is to perform alignment inference in the amino acid space
\parencite{bininda2005transalign,abascal2010translatorx}.
While this approach is an improvement over DNA models, it discards information,
underperforms compared to alignment at the codon level, and fails in the
presence of artifacts such as frameshifts and early stop codons.
% to correct this common practice is to discard lots of data
Given that errors and artifacts are common in molecular data, this requires
costly curation practices that discard large amounts of information.
% errors in genomic data that lead to erroneous downstream analyses
Errors in genomic datasets can lead to erroneous results in
functional and comparative genomic studies \parencite{estimates_schneider_2009}.

### 2022-08-15 pairwise_appnote.tex
Fix typo
Add title
Add lab comments

% title goes here:
\begin{flushleft}
{\Large\textbf{COATi: }}
% codon-aware pairwise aligner + artifacts
\newline
% authors go here:
\\
Juan J. Garcia Mesa\textsuperscript{1,2},
Reed A. Cartwright\textsuperscript{1,3}
\\
\bigskip
1 Biodesign Institute, Arizona State University
\\
2 Ira A. Schools of Engineering, Arizona State University
\\
3 School of Life Sciences, Arizona State University
\\
\bigskip
* corresponding@author.mail

@@

\section{Conclusions/Discussion}

COATi is an FST-based application that can find the optimal alignment between a
pair of sequences in the presence of artifacts using a statistical model.
% \green{(continue)}

@@

\subsection*{Funding}
This research was founded by an NSF-IIBR grant (\green{grant number}).\\

@@

% \section{References}
% \printbibliography

>>>>

% title goes here:
\begin{flushleft}
{\Large\textbf{COATi: statistical pairwise aligner robust to artifacts}}
% codon-aware pairwise aligner + artifacts
\newline
% authors go here:
\\
Juan J. Garcia Mesa\textsuperscript{1,2},
Reed A. Cartwright\textsuperscript{1,3}
\\
\bigskip
1 Biodesign Institute, Arizona State University
\\
2 Ira A. Schools of Engineering, Arizona State University
\\
3 School of Life Sciences, Arizona State University
\\
\bigskip
* corresponding@author.mail

@@

\section{Discussion}

COATi is an FST-based application that can calculate the optimal alignment
between a pair of sequences in the presence of artifacts using a statistical
model.
It will allow researchers to analyze more data with higher accuracy and
facilitate the study of important biological processes that shape genomic data.

@@

\subsection*{Funding}
This research was funded by an NSF-IIBR grant (\green{grant number}).\\

@@

% \section{References}
\printbibliography

### 2022-08-15 results.tex
Add lab comments

Using 4000 human genes and their gorilla homologous pairs from the ENSEMBL
database \parencite{ensembl_hubbard_2002} we simulated a data set of pairwise
alignments with empirical gap patterns.
We used the data set to evaluate the accuracy of popular cutting edge aligners
Clustal$\Omega$ v1.2.4 \parencite{clustal_omega_sievers_2011}, MACSE v2.06
\parencite{ranwez_macse_2011}, MAFFT v7.407 \parencite{mafft_katoh_2002}, and
PRANK v.170427 \parencite{prank_loytynoja_2014} together with COATi.

@@

COATi was significantly more accurate (lower $d_{seq}$) than other aligners; all
p-values were equal (MAFFT) or less than $1.714$x$10^{-8}$ according to the
one-tailed Wilcoxon signed rank test.
In addition, COATi produced more perfect alignments, less imperfect alignments,
and had a higher positive (97.3\%) selection accuracy.
MACSE (allows frameshifts) and MAFFT (using DNA) performed similarly with an
average alignment error an order of magnitude larger than COATi and the accuracy
retrieving positive selection of 81.5\% and 85.8\% respectively.
PRANK (using codons) and Clustal$\Omega$ (using amino acid translations) also
obtained similar results with an average alignment error two orders of magnitude
larger than COATi and a 87.3\% and 69.1\% accuracy retrieving positive selection.
The accuracy retrieving negative selection was similar across all five
aligners.

>>>>

Using 4000 human genes and their gorilla homologous pairs from the ENSEMBL
database \parencite{ensembl_hubbard_2002} we simulated a data set of pairwise
alignments with empirical gap patterns.
We used the data set to evaluate the accuracy of popular cutting edge aligners
Clustal$\Omega$ v1.2.4, %\parencite{clustal_omega_sievers_2011},
MACSE v2.06 \parencite{ranwez_macse_2011}, MAFFT v7.407, and
%\parencite{mafft_katoh_2002}
PRANK v.170427 % \parencite{prank_loytynoja_2014}
together with COATi.

@@

COATi was significantly (p < $1.714$x$10^{-8}$) more accurate at inferring
simulated alignments compared to other aligners according to the one-tailed
Wilcoxon signed ranked test, with an average alignment error ($d_{seq}$) value
of $6$x$10^{-4}$.
In addition, COATi produced more perfect alignments ($d_{seq}=0$), less imperfect
alignments, and more accurately retrieved events of positive selection (97.3\%).
Compared to COATi, MACSE (allows frameshifts) and MAFFT (using DNA) had an
average alignment error an order of magnitude larger than COATi ($6$x$10^{-3}$)
and a lower accuracy retrieving events of positive selection at a rate of 81.5\%
and 85.8\% respectively.
PRANK (using codons) and Clustal$\Omega$ (using amino acid translations) had
an average alignment error two orders of magnitude larger than COATi ($0.01$)
and a 87.3\% and 69.1\% accuracy retrieving events of positive selection,
respectively.
The accuracy retrieving events of negative selection was similar across all five
aligners (98.6\% $\pm$ 1.25\%).
